Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-04-01 doc/invt.tex
small edits

\begin{abstract}
  The Wang-Landau (WL) algorithm and metadynamics
  are two closely-related techniques
  in free energy calculations.
  %
  Both methods sample a flat distribution
  along a quantity of interest, $z$,
  by constructing a bias potential that offsets
  the potential of mean force (PMF) on the fly.
  %
  The two, however, differ by
  the manner of updating the bias potential:
  %
  the WL algorithm updates only
  the bin containing the current $z$,
  while metadynamics updates also
  several neighboring bins according to
  a Gaussian wave packet.
  %
  The error of the resulting PMF
  is determined by the updating schedule,
  or the updating magnitude
  as a function of simulation time.
  %
  In the WL case,
  the optimal schedule
  is known to be given by
  the inverse simulation time.
  %
  In this study,
  we give a formula of computing the optimal schedule
  for a general updating scheme
  involving multiple bins.
  %
  We show that while metadynamics can be advantageous
  for relatively short simulations,
  the single-bin updating scheme
  used in the WL algorithm
  belongs to a class of updating schemes
  optimal for asymptotic convergence.
  %the optimal schedule
  %and the effectiveness of metadynamics
  %are sensitive to the simulation length and other parameters.
  %
  %Further, we compared the two algorithms, and found that
  %although metadynamics produces a smoother PMF profile,
  %it is less effective in reducing errors with the shorter wavelengths.
  %%
  %Thus, the latter relies more heavily on the assumption of
  %the smoothness of the PMF profile,
  %especially with a wide updating window.
\end{abstract}

@@

In this study,
we present a method of computing
the optimal schedule
for a general updating scheme,
including those used in WL and metadynamics simulations.
%
The method reproduces the inverse-time formula
for the WL case.
%
It also predicts, for metadynamics,
a more algebraically complex schedule,
which is sensitive to the simulation length
and the width of the Gaussian wave packet.
%
Further, by comparing
different updating schemes,
we show that
the single-bin updating scheme
used by the WL algorithm
is best for infinitely long simulations.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.
%
The article is organized as follows.
%
We present the analytical results in Sec. \ref{sec:theory},
numerically verify some key aspects
in Sec. \ref{sec:results},
and conclude the article
in Sec. \ref{sec:conclusion}.

@@

Here we shall call a simulation that targets
a nearly flat distribution\cite{dayal2004, trebst2004, singh2011},
$p_i$,
a flat-distribution sampling (FDS) simulation.
%
Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
estimates the bias potential, $V_i$,
before simulation,
and fixes it during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{berg1992, lee1993}
and simulated tempering\cite{marinari1992, lyubartsev1992}
with $i$ being the index of energy
and that of temperature, respectively.
%
The estimated bias potential, $V_i$,
can often be improved
by the histogram along $i$ accumulated
during simulation
Using the histogram as $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1}
allows us to correct $V_i$ as
%
\begin{equation}
V^\mathrm{corr}_i
=
V_i
+
\log \frac{ \pi_i }
          {   p_i }.
\label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    \pi_i &\propto p^*_i \, e^{-V_i}, \\
    p_i &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
For an unknown system,
the main difficulty of the process
is that it is hard to guess a sufficiently accurate
$V_i$ to start with.
%
Thus,
it can take many iterative rounds of simulations
to achieve the desired flat distribution.
%
Further, a reasonably precise histogram
requires an accumulation period
longer than the autocorrelation time,
which can also be hard to estimate beforehand.

@@

We first approximate Eq. \eqref{eq:wl_update}
by a differential equation
%
\begin{equation}
  \dot v_i(t)
  =
  \alpha(t) \, \frac{ h_i(t) } { p_i },
  \label{eq:vt_diffeq}
\end{equation}
%
where
$\dot v_i(t) \equiv dv_i(t)/dt$,
%
and $h_i(t) = \delta_{i, i(t)}$
is the instantaneous histogram,
which is equal to $1.0$
for the current bin $i(t)$
or zero otherwise.

@@

Now by diagonalizing the matrix $\mathbf w$
using the eigenvectors defined in Eqs. \eqref{eq:eig_w},
\eqref{eq:eig_orthonormal_cols},
and
\eqref{eq:eig_orthonormal_rows},
we get a set of decoupled equations
%
\begin{equation}
\dot y_k(t)
=
-\alpha(t) \, \lambda_k \, [y_k(t) - \eta_k(t)],
\notag
%\label{eq:yt_diffeq}
\end{equation}
%
where
\begin{align}
  y_k &= \sum_{j=1}^n \phi_{kj} \, x_j,
  \label{eq:y_def}
  \\
  \eta_k &= \sum_{j=1}^n \phi_{kj} \frac{ \zeta_j}{ p_j}.
  \label{eq:eta_def}
\end{align}

@@

In analogous to Eq. \eqref{eq:x2t_average},
we have, from Eq. \eqref{eq:y2_sum},
the error after a period of $T$,
%
\begin{equation}
  \Err
  =
  \Err_R + \Err_A,
  \label{eq:error_split}
\end{equation}
%
which is the sum of the residual error
%
\begin{equation}
  \Err_R
  =
  \sum_{k = 2}^n
    \left\langle y_k^2(0) \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(T) },
  \label{eq:error_res}
\end{equation}
%
from the decay of the initial error,
and the asymptotic error
\begin{equation}
  \Err_A
  =
  \int_0^T \int_0^T
  \sum_{k = 2}^n
    \dot u_k\bigl( q(t) \bigr) \,
    \dot u_k\bigl( q(t') \bigr) \,
    \kappa_k(t - t') \, dt \, dt',
  \label{eq:error_asym}
\end{equation}
%
where,
\begin{align*}
  u_k(q')
  &\equiv
  e^{\lambda_k \, [q' - q(T)]},
  \\
  \kappa_k(t - t')
  &\equiv
  \left\langle
    \eta_k(t) \, \eta_k(t')
  \right\rangle.
\end{align*}
%
\note{The extremal condition of $\Err_A$ is given by
%
\begin{align}
\sum_{k = 2}^n
\lambda_k \, u_k\bigl( q(t) \bigr)
\int_0^T
  \ddot u_k \bigl( q(t') \bigr) \,
  \kappa_k(t' - t) \, dt' = 0.
\label{eq:optimal_mbin}
\end{align}
%
If all $\lambda_k$, hence all $u_k$, are the same,
the above condition can be satisfied
by Eq. \eqref{eq:ddu_eq_0},
which is the single-bin case.
%
Unfortunately, the above expression
appears to be useless in a general setting.
}

@@

As shown above, with fixed endpoint value, $q(T)$,
the optimization of the total error is equivalent to
that of the asymptotic part.
%
We can rewrite Eq. \eqref{eq:error_asym1} as an action
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt,
  \notag
  %\label{eq:error_asym_action}
\end{equation}
%
where the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ),
  \label{eq:error_asym_lagr}
\end{align}
%
represents a free particle
with a position-dependent mass.
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian,
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \partial \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
which coincides with the Lagrangian,
is conserved.
%
So we may set
%
\begin{equation}
  \mathcal L
  =
    \textstyle\sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  c^2
  \quad
  (c > 0).
  \label{eq:Lagrangian_const}
\end{equation}
%
Integrating the square root of the left-hand side, we get
%
\begin{equation}
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  c \, t
  ,
  \label{eq:q_opt}
\end{equation}
%
where $c$ is to be determined by
the boundary condition at $t = T$ as
%
\begin{equation}
  c =
  \frac 1 T
  \int_{ 0 }^{ q(T) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  .
  \notag
  %\label{eq:q_opt}
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.

@@

@@

\subsubsection{\label{sec:bandkernel}
Window kernel and eigenvalues}

It is natural to introduce a homogeneous updating scheme
for a periodic variable\cite{dama2014}.
%
Consider the following matrix, $\mathbf w$,
%
\begin{equation}
  w_{ij}
  =
  \mu_{i-j}
  +
  \mu_{i-j+n}
  +
  \mu_{i-j-n}
  ,
\label{eq:w_band_pbc}
\end{equation}
%
where
$\mu_{-b}, \dots, \mu_b$ ($b < n/2$)
are a set of numbers satisfying
%
\begin{equation}
  \mu_{-b} + \cdots + \mu_b = 1
  ,
\label{eq:mu_normalization}
\end{equation}
%
and $\mu_l = 0$ for $l > b$.
%
Particularly,
we shall only consider symmetric kernels
that satisfy
%
\begin{equation}
  \mu_i = \mu_{-i}
  .
\label{eq:mu_symm}
\end{equation}
%
These numbers constitute a \emph{updating kernel}.

@@

Since $\lambda_1$ corresponds to the mode
of a uniform shift of all $v_k$,
we can freely set $\lambda_1$ to $\lambda_2$
without changing the nature of the updating scheme.
%
Now an updating matrix $\mathbf w$
with identical eigenvalues
is essentially a multiple of the identity matrix.
%
Thus, in terms of asymptotic convergence,
the single-bin scheme
(as adopted in the WL algorithm)
is most efficient among updating schemes.
%
In the single-bin case,
$c_k = 1/(T + t_0)$,
and the error is
\begin{align}
  E
  =
  \frac{       T     }
       { (T + t_0)^2 }
  \sum_{ k = 2 }^n
    \Gamma_k
  .
\label{eq:error_singlebin}
\end{align}

@@

@@

@@

A practical question now arises.
%
In late stages of an adaptive FDS simulation,
if we are confident that the bias potential
is sufficiently accurate,
will we be better off switching to the equilibrium FDS method
to prevent possible efficiency loss
due to the adaptive updates?
%
Below, we shall show that this is unnecessary
if the simulation uses the single-bin (WL) updating scheme
with the optimal schedule, Eq. \eqref{eq:alpha_invt}.
%
That is, the bias potential obtained from
an optimal single-bin scheme FDS simulation
will be similar in precision to that from
the ideal equilibrium FDS simulation
in long times.
%
Thus, our demonstration below serves as an alternative proof
of the optimality of the single-bin scheme.

The error of the equilibrium FDS method
depends on the precision of the histogram.
%
The error is minimized
when $v_i$'s assume the exact values,
%
and in this case,
the error came from the second term
on the right-hand side of Eq. \eqref{eq:vcorr_equil},
%
\begin{align}
  E
  &=
  \left\langle
    \sum_{ i = 1 }^ n
      p_i \,
      \left(
        \log \frac { \bar h_i }
                   {      p_i }
      \right)^2
  \right\rangle
\notag
\\
  &\approx
  \sum_{ i = 1 }^ n
    p_i \,
    \left\langle
      \left(
        \frac { \bar h_i }
              {      p_i }
        - 1
      \right)^2
    \right\rangle.
\notag
\end{align}

To make better comparison with
the error of the single-bin case,
we shall rewrite the error as
a sum over the eigenmodes.
We define
\begin{align}
  \bar \eta_k
  =
  \frac{ 1 } { T }
  \sum_{ t = 1 } ^ T
    \eta_k( t )
  ,
\notag
\end{align}
%
with $\eta_k(t_1)$ defined by Eq. \eqref{eq:eta_def}
and $\phi_{k i}$ by Eq. \eqref{eq:eig_orthonormal_cols}.
%\begin{align}
%\eta_k( t_1 )
%=
%\sum_{ i = 1 }^ n
%\phi_{k i} \left[
%             \frac{ h_i(t_1) }
%                  { p_i      }
%             - 1
%           \right],
%\end{align}
%and
%\begin{align}
%\sum_{ k = 1 }^n
%  \phi_{k i} \, \phi_{k j}
%  = p_i \, \delta_{i j}.
%\end{align}
%
Then,
%
\begin{align}
  E
  =
  \sum_{ k = 1 }^n
    \left\langle
      \bar \eta_k^2
    \right\rangle.
\notag
%\label{eq:error_sbin_equil}
\end{align}
%
\note{Because
%
\begin{align*}
\sum_{ k = 1 }^n
  \bar \eta_k^2
&=
\frac{ 1 } { T^2 }
\sum_{ t, t' = 1 }^T
  \sum_{ k = 1 }^n
    \eta_k( t ) \, \eta_k( t' )
\\
&=
\frac{ 1 } { T^2 }
\sum_{ t, t' = 1 }^T
  \sum_{ i, j = 1 }^n
    \left[
      \frac{ h_i(t) }
           { p_i    }
      - 1
    \right]
    \left[
      \frac{ h_j(t') }
           { p_j     }
      - 1
    \right]
    \sum_{ k = 1 }^n
      \phi_{k \, i} \, \phi_{k \, j}
\\
&=
\frac{ 1 } { T^2 }
\sum_{ t, t' = 1 }^T
  \sum_{ i, j = 1 }^n
    \left[
      \frac{ h_i(t) }
           { p_i    }
      - 1
    \right]
    \left[
      \frac{ h_j(t') }
           { p_j     }
      - 1
    \right]
    p_i \, \delta_{i, j}
\\
&=
\sum_{ i = 1 }^n
\frac{ 1 } { T }
\sum_{ t = 1 }^T
    \left[
      \frac{ h_i(t) }
           { p_i    }
      - 1
    \right]
\frac{ 1 } { T }
\sum_{ t' = 1 }^T
    \left[
      \frac{ h_i(t') }
           { p_i     }
      - 1
    \right]
\\
&=
\sum_{ i = 1 }^n
  p_i \,
    \left(
      \frac{ \bar h_i }
           { p_i      }
      - 1
    \right)^2
.
\end{align*}
}
The sum can start from $k = 2$, for
by Eq. \eqref{eq:eigenmode1}, we have
$$
\eta_1(t)
=
\sum_{ i = 1 }^n
  \phi_{1i} \, \frac{ h_i(t) - p_i } { p_i }
=
\sum_{ i = 1 }^n h_i(t)
-
\sum_{ i = 1 }^n p_i = 0.
$$

In the long-time limit,
\begin{align*}
\left\langle
  \bar \eta_k^2
\right\rangle
&=
\frac{1}{T^2}
\sum_{ t, t' = 1 }^T
\langle \eta_k(t) \, \eta_k(t') \rangle
\\
&\approx
\frac{1}{T}
\sum_{ \tau = -\infty }^{ \infty }
\langle \eta_k(t) \, \eta_k(t + \tau) \rangle
\\
&=
\frac{1}{T}
\int_{-\infty}^\infty
\kappa_n(\tau) \, d\tau
=
\frac{1}{T}
\Gamma_k.
\end{align*}
%
Thus,
\begin{equation}
  E
  =
  \frac{ 1 } { T }
  \sum_{ k = 2 }^n \Gamma_k
  .
\label{eq:error_equil}
\end{equation}

@@

For the single-bin updating scheme,
we chose $\alpha_0 = 0.0001$
and ran the simulation for $t = 10^8$ steps.
%
As shown in Fig. \ref{fig:err_singlebin},
the numerical results of the terminal error agreed well with
the analytical prediction by Eq. \eqref{eq:error_invt}.
%
For both global and local sampling processes,
the optimal $\lambda$ occurred around $1.0$,
verifying Eq. \eqref{eq:alpha_invt}.
%
%Recall that $1/\lambda$ in Eq. \eqref{eq:alpha_invtlambda}
%represents the relative updating magnitude.
%%
%Figure \ref{fig:err_singlebin} also shows
%an asymmetry of the error curve.
%A smaller than optimal value of
%the updating magnitude, $1/\lambda$,
%tends to increase the error more rapidly
%than a larger than optimal value,
%suggesting that the over-updating
%is generally preferable to the under-updating
%when the optimal value is unknown.

>>>>

\begin{abstract}
  The Wang-Landau (WL) algorithm and metadynamics
  are two closely-related techniques
  in free energy calculations.
  %
  Both methods sample a flat distribution
  along a quantity of interest, $z$,
  by constructing a bias potential that offsets
  the potential of mean force (PMF) on the fly.
  %
  The two, however, differ by
  the manner of updating the bias potential:
  %
  the WL algorithm updates only
  the bin containing the current $z$,
  while metadynamics updates also
  several neighboring bins according to
  a Gaussian wave packet.
  %
  The error of the resulting PMF
  is determined by the updating schedule,
  or the updating magnitude
  as a function of simulation time.
  %
  In the WL case,
  the optimal schedule
  is known to be given by
  the inverse simulation time.
  %
  In this study,
  we give a method of computing the optimal schedule
  for a general updating scheme
  involving multiple bins.
  %
  We show that the optimal schedule of metadynamics
  depends on the simulation length
  and the width of the Gaussian.
  %is solved from the equation of motion
  %of a free particle with a position-dependent mass.
  %
  Further,
  the single-bin updating scheme
  used in the WL algorithm
  belongs to a class of multiple-bin updating schemes
  optimal for asymptotic convergence,
  and the optimal schedules of these schemes
  are all given by the inverse-time formula.
\end{abstract}

@@

In this study,
we present a method of computing
the optimal schedule
for a general updating scheme,
including those used in WL and metadynamics simulations.
%
The method reproduces the inverse-time formula
for the WL case.
%
It also predicts, for metadynamics,
a more algebraically complex schedule,
which is sensitive to the simulation length
and the width of the Gaussian wave packet.
%
Further, by comparing
different updating schemes,
we show that
the single-bin updating scheme
used by the WL algorithm
can be generalized to class of multiple-bin updating schemes
that are partially optimal for infinitely long simulations,
and the inverse-time schedule is optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.
%
The article is organized as follows.
%
We present the analytical results in Sec. \ref{sec:theory},
numerically verify some key aspects
in Sec. \ref{sec:results},
and conclude the article
in Sec. \ref{sec:conclusion}.

@@

Here we shall call a simulation that targets
a nearly flat distribution\cite{dayal2004, trebst2004, singh2011},
$p_i$,
a flat-distribution sampling (FDS) simulation.
%
Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
estimates the bias potential, $V_i$,
before simulation,
and fixes it during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{berg1992, lee1993}
and simulated tempering\cite{marinari1992, lyubartsev1992}
with $i$ being the index of energy
and that of temperature, respectively.
%
The estimated bias potential, $V_i$,
can often be improved
by the histogram along $i$ accumulated
during simulation
Using the histogram $H_i$ as $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1}
allows us to correct $V_i$ as
%
\begin{equation}
V^\mathrm{corr}_i
=
V_i
+
\log \frac{ H_i }
          { p_i }.
\label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    \pi_i &\propto p^*_i \, e^{-V_i}, \\
    p_i &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
For an unknown system,
the main difficulty of the process
is that it is hard to guess a sufficiently accurate
$V_i$ to start with.
%
Thus,
it can take many iterative rounds of simulations
to achieve the desired flat distribution.
%
Further, a reasonably precise histogram
requires an accumulation period
longer than the autocorrelation time,
which can also be hard to estimate beforehand.

@@

We first approximate Eq. \eqref{eq:wl_update}
by a differential equation
%
\begin{equation}
  \dot v_i(t)
  =
  \alpha(t) \, \frac{ h_i(t) } { p_i },
  \label{eq:vt_diffeq}
\end{equation}
%
where
$\dot v_i(t) \equiv dv_i(t)/dt$,
%
and
\begin{equation}
  h_i(t) = \delta_{i, i(t)}
  ,
  \label{eq:h_def}
\end{equation}
is the instantaneous histogram,
which is equal to $1.0$
for the current bin $i(t)$
or zero otherwise.

@@

Now by diagonalizing the matrix $\mathbf w$
using the eigenvectors defined in Eqs. \eqref{eq:eig_w},
\eqref{eq:eig_orthonormal_cols},
and
\eqref{eq:eig_orthonormal_rows},
we get a set of decoupled equations
%
\begin{equation}
\dot y_k(t)
=
-\alpha(t) \, \lambda_k \, [y_k(t) - \theta_k(t)],
\notag
%\label{eq:yt_diffeq}
\end{equation}
%
where
\begin{align}
  y_k &= \sum_{j=1}^n \phi_{kj} \, x_j,
  \label{eq:y_def}
  \\
  \theta_k &= \sum_{j=1}^n \phi_{kj} \frac{ \zeta_j}{ p_j}.
  \label{eq:eta_def}
\end{align}

@@

In analogous to Eq. \eqref{eq:x2t_average},
we have, from Eq. \eqref{eq:y2_sum},
the error after a period of $T$,
%
\begin{equation}
  \Err
  =
  \Err_R + \Err_A,
  \label{eq:error_split}
\end{equation}
%
which is the sum of the residual error
%
\begin{equation}
  \Err_R
  =
  \sum_{k = 2}^n
    \left\langle y_k^2(0) \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(T) },
  \label{eq:error_res}
\end{equation}
%
from the decay of the initial error,
and the asymptotic error
\begin{equation}
  \Err_A
  =
  \int_0^T \int_0^T
  \sum_{k = 2}^n
    \dot u_k\bigl( q(t) \bigr) \,
    \dot u_k\bigl( q(t') \bigr) \,
    \kappa_k(t - t') \, dt \, dt',
  \label{eq:error_asym}
\end{equation}
%
where,
\begin{align*}
  u_k(q')
  &\equiv
  e^{\lambda_k \, [q' - q(T)]},
  \\
  \kappa_k(t - t')
  &\equiv
  \left\langle
    \theta_k(t) \, \theta_k(t')
  \right\rangle.
\end{align*}
%
\note{The extremal condition of $\Err_A$ is given by
%
\begin{align}
\sum_{k = 2}^n
\lambda_k \, u_k\bigl( q(t) \bigr)
\int_0^T
  \ddot u_k \bigl( q(t') \bigr) \,
  \kappa_k(t' - t) \, dt' = 0.
\label{eq:optimal_mbin}
\end{align}
%
If all $\lambda_k$, hence all $u_k$, are the same,
the above condition can be satisfied
by Eq. \eqref{eq:ddu_eq_0},
which is the single-bin case.
%
Unfortunately, the above expression
appears to be useless in a general setting.
}

@@

As shown above, with fixed endpoint value, $q(T)$,
the optimization of the total error is equivalent to
that of the asymptotic part.
%
We can rewrite Eq. \eqref{eq:error_asym1} as an action
%
\begin{equation}
  \Err_A
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt,
  \notag
  %\label{eq:error_asym_action}
\end{equation}
%
where the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl[ q(t) \bigr]
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl[ q(t) \bigr]
  \; \dot q^2( t ),
  \label{eq:error_asym_Lagrangian}
\end{align}
%
represents a free particle
with a position-dependent mass.
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian,
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \partial \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
which coincides with the Lagrangian,
is conserved.
%
So we may set
%
\begin{equation}
  \mathcal L
  =
    \textstyle\sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(t) \bigr]
  \;
  \dot q^2(t)
  =
  c^2
  \quad
  (c > 0).
  \label{eq:Lagrangian_const}
\end{equation}
%
Integrating the square root of the left-hand side, we get
%
\begin{equation}
  \int_{ 0 }^{ q(t) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  c \, t
  ,
  \label{eq:q_opt}
\end{equation}
%
where $c$ is to be determined by
the boundary condition at $t = T$ as
%
\begin{equation}
  c =
  \frac 1 T
  \int_{ 0 }^{ q(T) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  .
  \notag
  %\label{eq:q_opt}
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.

@@

In practice, Eq. \eqref{eq:varY}
may fail if some eigenvalue $\lambda_k$ is zero.
%
However, the corresponding modes can be ignored
as they do not contribute to the Lagrangian defined by
Eq. \eqref{eq:error_asym_Lagrangian}.

@@

\subsubsection{\label{sec:bandkernel}
Updating kernel and eigenvalues}

It is natural to introduce a homogeneous updating scheme
for a periodic variable\cite{dama2014}.
%
Consider the following matrix, $\mathbf w$,
%
\begin{equation}
  w_{ij}
  =
  \mu_{i-j}
  +
  \mu_{i-j+n}
  +
  \mu_{i-j-n}
  ,
\label{eq:w_band_pbc}
\end{equation}
%
where
$\mu_{-b}, \dots, \mu_b$ ($b \le n/2$)
are a set of numbers satisfying
%
\begin{equation}
  \mu_{-b} + \cdots + \mu_b = 1
  ,
\label{eq:mu_normalization}
\end{equation}
%
and $\mu_l = 0$ for $l > b$.
%
Particularly,
we shall only consider symmetric kernels
that satisfy
%
\begin{equation}
  \mu_i = \mu_{-i}
  .
\label{eq:mu_symm}
\end{equation}
%
If $b = n/2$, then $\mu_{-b}$ is dropped
from Eq. \eqref{eq:normalization}.
%
These numbers constitute a \emph{updating kernel}.

@@

Since $\lambda_1$ corresponds to the mode
of a uniform shift of all $v_k$,
we can freely set $\lambda_1$ to $\lambda_2$
without changing the nature of the updating scheme.
%
Now an updating matrix $\mathbf w$
with identical eigenvalues
is essentially a multiple of the identity matrix.
%
Thus, if no eigenvalue $\lambda_k$ is zero,
then in terms of asymptotic convergence,
the single-bin scheme adopted in the WL algorithm
is most efficient.
%
In the single-bin case,
$c_k = 1/(T + t_0)$,
and the error is
\begin{align}
  E
  =
  \frac{       T     }
       { (T + t_0)^2 }
  \sum_{ k = 2 }^n
    \Gamma_k
  .
\label{eq:error_singlebin}
\end{align}

@@

\subsubsection{\label{sec:optscheme}
Optimal schemes}

@@

The optimal schedule of these schemes
remains to be the inverse-time schedule,
given by Eq. \eqref{eq:alpha_invt1}.

@@

A practical question now arises.
%
In late stages of an adaptive FDS simulation,
if we are confident that the bias potential
is sufficiently accurate,
will we be better off switching to the equilibrium FDS method
to prevent possible efficiency loss
due to the adaptive updates?
%
Below, we shall show that this is unnecessary
if the simulation uses the single-bin (WL) updating scheme
with the optimal schedule, Eq. \eqref{eq:alpha_invt1}.
%
That is, the bias potential obtained from
an optimal single-bin scheme FDS simulation
will be similar in precision to that from
the ideal equilibrium FDS simulation
in long times.
%
Thus, our demonstration below serves as an alternative proof
of the optimality of the single-bin scheme.

The error of the equilibrium FDS method
depends on the precision of the histogram.
%
The error is minimized
when $v_i$'s assume the exact values,
%
and in this case,
the error came from the second term
on the right-hand side of Eq. \eqref{eq:vcorr_equil},
%
\begin{align}
  E
  &=
  \left\langle
    \sum_{ i = 1 }^ n
      p_i \,
      \left(
        \log \frac { H_i }
                   { p_i }
      \right)^2
  \right\rangle
\notag
\\
  &\approx
  \sum_{ i = 1 }^ n
    p_i \,
    \left\langle
      \left(
        \frac { H_i }
              { p_i }
        - 1
      \right)^2
    \right\rangle,
\notag
\end{align}
where
$H_i = \frac{1}{T} \sum_{t = 1}^T h_i(t)$
with
$h_i(t)$ defined by Eq. \eqref{eq:h_def}.

To make better comparison with
the error of the single-bin case,
we shall rewrite the error as
a sum over the eigenmodes.
We define
\begin{align}
  \Theta_k
  =
  \frac{ 1 } { T }
  \sum_{ t = 1 } ^ T
    \theta_k( t )
  ,
\notag
\end{align}
%
with $\theta_k(t_1)$ defined by Eq. \eqref{eq:eta_def}
and $\phi_{k i}$ by Eq. \eqref{eq:eig_orthonormal_cols}.
%\begin{align}
%\theta_k( t_1 )
%=
%\sum_{ i = 1 }^ n
%\phi_{k i} \left[
%             \frac{ h_i(t_1) }
%                  { p_i      }
%             - 1
%           \right],
%\end{align}
%and
%\begin{align}
%\sum_{ k = 1 }^n
%  \phi_{k i} \, \phi_{k j}
%  = p_i \, \delta_{i j}.
%\end{align}
%
Then,
%
\begin{align}
  E
  =
  \sum_{ k = 1 }^n
    \left\langle
      \Theta_k^2
    \right\rangle.
\notag
%\label{eq:error_sbin_equil}
\end{align}
%
\note{Because
%
\begin{align*}
\sum_{ k = 1 }^n
  \Theta_k^2
&=
\frac{ 1 } { T^2 }
\sum_{ t, t' = 1 }^T
  \sum_{ k = 1 }^n
    \theta_k( t ) \, \theta_k( t' )
\\
&=
\frac{ 1 } { T^2 }
\sum_{ t, t' = 1 }^T
  \sum_{ i, j = 1 }^n
    \left[
      \frac{ h_i(t) }
           { p_i    }
      - 1
    \right]
    \left[
      \frac{ h_j(t') }
           { p_j     }
      - 1
    \right]
    \sum_{ k = 1 }^n
      \phi_{k \, i} \, \phi_{k \, j}
\\
&=
\frac{ 1 } { T^2 }
\sum_{ t, t' = 1 }^T
  \sum_{ i, j = 1 }^n
    \left[
      \frac{ h_i(t) }
           { p_i    }
      - 1
    \right]
    \left[
      \frac{ h_j(t') }
           { p_j     }
      - 1
    \right]
    p_i \, \delta_{i, j}
\\
&=
\sum_{ i = 1 }^n
\frac{ 1 } { T }
\sum_{ t = 1 }^T
    \left[
      \frac{ h_i(t) }
           { p_i    }
      - 1
    \right]
\frac{ 1 } { T }
\sum_{ t' = 1 }^T
    \left[
      \frac{ h_i(t') }
           { p_i     }
      - 1
    \right]
\\
&=
\sum_{ i = 1 }^n
  p_i \,
    \left(
      \frac{ \bar h_i }
           { p_i      }
      - 1
    \right)^2
.
\end{align*}
}
The sum can start from $k = 2$, for
by Eq. \eqref{eq:eigenmode1}, we have
$$
\theta_1(t)
=
\sum_{ i = 1 }^n
  \phi_{1i} \, \frac{ h_i(t) - p_i } { p_i }
=
\sum_{ i = 1 }^n h_i(t)
-
\sum_{ i = 1 }^n p_i = 0.
$$

In the long-time limit,
\begin{align*}
\left\langle
  \Theta_k^2
\right\rangle
&=
\frac{1}{T^2}
\sum_{ t, t' = 1 }^T
\langle \theta_k(t) \, \theta_k(t') \rangle
\\
&\approx
\frac{1}{T}
\sum_{ \tau = -\infty }^{ \infty }
\langle \theta_k(t) \, \theta_k(t + \tau) \rangle
\\
&=
\frac{1}{T}
\int_{-\infty}^\infty
\kappa_n(\tau) \, d\tau
=
\frac{1}{T}
\Gamma_k.
\end{align*}
%
Thus,
\begin{equation}
  E
  =
  \frac{ 1 } { T }
  \sum_{ k = 2 }^n \Gamma_k
  .
\label{eq:error_equil}
\end{equation}

@@

For the single-bin updating scheme,
we chose $\alpha_0 = 0.0001$
and ran the simulation for $t = 10^8$ steps.
%
As shown in Fig. \ref{fig:err_singlebin},
the numerical results of the terminal error agreed well with
the analytical prediction by Eq. \eqref{eq:error_invt}.
%
For both global and local sampling processes,
the optimal $\lambda$ occurred around $1.0$,
verifying Eq. \eqref{eq:alpha_invt1}.
%
%Recall that $1/\lambda$ in Eq. \eqref{eq:alpha_invtlambda}
%represents the relative updating magnitude.
%%
%Figure \ref{fig:err_singlebin} also shows
%an asymmetry of the error curve.
%A smaller than optimal value of
%the updating magnitude, $1/\lambda$,
%tends to increase the error more rapidly
%than a larger than optimal value,
%suggesting that the over-updating
%is generally preferable to the under-updating
%when the optimal value is unknown.
